The Concurrent Calculi Formalisation Benchmark

### 2024-02-17 short-paper/short-paper.tex
some rephrasing of initial paragraphs in the intro
some comments on the general organization

cleanup of duplication doi/url in bib

% - Introduction
%   - POPLMark accelerated progress in mechanisation tools for PL
%     - POPLMark Reloaded tried to do the same for logical relations
%     - But they were deliberately limited in scope to not include process calculi
%   - We propose a new collection of benchmark problems for process calculi
%     - Why not concurrency in general? (after this point, only mention process calculi)
%     - Why is there a problem with process calculus mechanisations?
%     - Our experience says there is, and we will cite other experiences below
%   - We have noticed three key issues
%     - They are not covered by POPLMark
%     - Techniques are less settled than for binders
%     - Others also mention them as problems: cite and explain references briefly for each challenge topic
%   - Our goals are to determine the state of the art, develop best practices and tutorials, and identify improvements to proof assistant technology that would materially help the situation
\section{Introduction}
The POPLMark challenge~\cite{POPLMark} accelerated the development of proof assistants, libraries, and best practices for programming language mechanisation by establishing a common ground for systematically evaluating mechanisation techniques.
In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs involving logical relations.
These challenges, and the developments they spurred, were limited in scope, but their authors noted the need to tackle topics such as coinduction and linearity in the future.

In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising formal models of concurrent and distributed programming languages in general, and process calculi in particular.
Concurrent calculi are notably subtle: for instance it took nine years before the incorrect proof of type safety in the seminal paper~\cite{Honda1998} was fixed in~\cite{Yoshida2007}.
Similarly, many key results in peer-reviewed papers on session types have turned out to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the value of machine-checkable proofs.
While results about concurrent formalisms have already been mechanised (as we will discuss further below), our experience is that choosing appropriate mechanisation techniques and tools remains a significant challenge and that the trade-offs are not well understood.
This often leads researchers toward a trial-and-error approach, resulting in sub-optimal solutions, wasted mechanisation effort, and techniques and results that are hard to reuse.
\cite{Cruz-Filipe2021b} notes that the high-level parts of mechanised proofs closely resemble the informal versions, and that the main challenge lies in getting the foundations right.

Our benchmark problems (explained in detail in \cref{app:challenges}) focus on three key aspects which often cause difficulties when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, and \emph{coinductive reasoning}.
Mechanisations must often address several of these aspects, but we see the combination of techniques as a next step beyond the scope of this challenge, as discussed further in \Cref{sec:going-beyond}.
Our benchmark challenges are based on process calculi, as these provide a simple but realistic exhibit of typical issues encountered in mechanisation work.
The three aspects are certainly not the only causes of difficulties, but as we will discuss presently, they emerge in most mechanisations of concurrency theory.

\todo{These paragraphs are still quite messy and unfocused. They should focus on the motivation, i.e.\ mention what others have perceived as problems.}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

@@

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.\alb{Do we need all those citations about encodings?}
Still, these (and other) applications need the basic techniques that our challenge problems emphasize: as per design goal \ref{item:goal-reusability}, our problems are drawn from the basic meta-theory of process calculi (without requiring the development of new theory), and focus on well-known theorems and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

@@

\lore{Need to integrate the following with the above.}
Different process-calculi
mechanisations
implement different techinques;
most of the time these 
are ad-hoc efforts to achieve a specific result.
\lore{cite and substantiate.}
In parallel, although most proof assistants
offer support for coinduction in their standard libraries,
research has mainly focused on the foundations
of the specific system, more than on
applicability.
\lore{cite and substantiate.}

>>>>

% - Introduction
%   - POPLMark accelerated progress in mechanisation tools for PL
%     - POPLMark Reloaded tried to do the same for logical relations
%     - But they were deliberately limited in scope to not include process calculi
%   - We propose a new collection of benchmark problems for process calculi
%     - Why not concurrency in general? (after this point, only mention process calculi)
%     - Why is there a problem with process calculus mechanisations?
%     - Our experience says there is, and we will cite other experiences below
%   - We have noticed three key issues
%     - They are not covered by POPLMark
%     - Techniques are less settled than for binders
%     - Others also mention them as problems: cite and explain references briefly for each challenge topic
%   - Our goals are to determine the state of the art, develop best practices and tutorials, and identify improvements to proof assistant technology that would materially help the situation
\section{Introduction}

The POPLMark challenge~\cite{POPLMark}
played a pivotal role in advancing the field of proof assistants,
libraries, and best practices for programming language
mechanization. By providing a shared framework for systematically
evaluating mechanization techniques, it catalyzed a significant
shift within the programming language research community. As a
result, scholarly publications increasingly incorporate
mechanized proofs.
%
% The POPLMark challenge~\cite{POPLMark} accelerated the development of proof assistants, libraries, and best practices for programming language mechanisation by establishing a common ground for systematically evaluating mechanisation techniques.
% In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
% POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs involving logical relations.
% These challenges, and the developments they spurred, were limited in scope, but their authors noted the need to tackle topics such as coinduction and linearity in the future.

POPLMark Reloaded~\cite{POPLMarkReloaded} introduced a program akin to
the original POPLMark challenge, specifically targeting proofs related
to logical relations. While these initiatives had a narrow focus,
their authors recognized the importance of addressing broader topics
in the future, including coinduction and linearity.

In this spirit, we propose a new collection of benchmark problems
specifically designed to address typical issues that arise when
mechanising formal models of concurrent and distributed programming
languages in general, and process calculi in particular.  Concurrent
calculi are notably subtle: for instance it took nine years before the
incorrect proof of type safety in the seminal paper by Honda et
al.~\cite{Honda1998} was fixed in~\cite{Yoshida2007}.  Similarly, many
key results in peer-reviewed papers on session types have turned out
to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the
value of machine-checkable proofs.

While results about concurrent
formalisms have already been mechanised (as we will discuss further
below), our experience is that choosing appropriate mechanisation
techniques and tools remains a significant challenge and that the
trade-offs are not well understood.  This often leads researchers
toward a trial-and-error approach, resulting in sub-optimal solutions,
wasted mechanisation effort, and techniques and results that are hard
to reuse.  For example, the authors in~\cite{Cruz-Filipe2021b} note that the high-level parts of
mechanised proofs closely resemble the informal versions, and that the
main challenge lies in getting the foundations right.

Our benchmark problems (explained in detail in \cref{app:challenges}) center around three critical aspects that frequently pose challenges in mechanizing concurrency theory:
% focus on three key aspects which often cause difficulties when mechanising concurrency theory:
\emph{linearity}, \emph{scope extrusion}, and \emph{coinductive
  reasoning}.  Mechanisations must often address several of these
aspects at the same time; however, we see the combination of
techniques as a next step beyond the scope of this challenge, as
discussed further in \Cref{sec:going-beyond}.  Our benchmark
challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanization endeavors
% exhibit of typical issues encountered in mechanisation work.
While the three highlighted aspects are not the sole sources of difficulties, they
consistently manifest in most mechanizations of concurrency theory.
% The three aspects are certainly not the only causes of difficulties,
% but as we will discuss presently, they emerge in most mechanisations
% of concurrency theory.

\todo{These paragraphs are still quite messy and unfocused. They should focus on the motivation, i.e.\ mention what others have perceived as problems.}

\alb{I really do not like the duplication of the discussion of the
  three issues here and in Sec 2. I understand the desire to highlight
  noted issues, but it does not work. The citations are unconvincing, furthermore}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

@@

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is
not meant to be comprehensive: applications such as multiparty session
types~\cite{10.1145/2827695,10.1145/3290343},
choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation
types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696},
or encodings between different
calculi~\cite{DBLP:journals/iandc/Gorla10,%,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,
  DBLP:conf/ecoop/ScalasDHY17}
  %,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407}
are not directly covered.

\alb{I've commented some citations about
  encodings. Is it a good idea to start the section describing what we \textbf{do not do}?}

Still, these (and other) applications need the basic
techniques that our challenge problems emphasize: as per design goal
\ref{item:goal-reusability}, our problems are drawn from the basic
meta-theory of process calculi (without requiring the development of
new theory), and focus on well-known theorems and results that involve
interesting proof techniques that may be reused in further work.  To
achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem should be
reasonably small and easily understood with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

@@

\lore{Need to integrate the following with the above.}
Different process-calculi
mechanisations
implement different techniques;
most of the time these 
are ad-hoc efforts to achieve a specific result.
\lore{cite and substantiate.}
In parallel, although most proof assistants
offer support for coinduction in their standard libraries,
research has mainly focused on the foundations
of the specific system, more than on
applicability.
\lore{cite and substantiate.}
